A Way Forward in Parallelising Dynamic Languages

### 2014-06-17 talk/icooolps2014/position-paper.tex
add copyright data
first round of integrating feedback

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

@@

@@

\conferenceinfo{ICOOOLPS workshop 2014}{July 28th, 2014, Uppsala, Sweden}
\copyrightyear{2014}
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

@@

These requirements are not easy to meet. We argue that STM is the
overall winner. While it has a big performance problem currently, it
gets more points in all the other categories. We think that it is the
only solution that also provides a better synchronisation mechanism to
the application in the form of parallelisable atomic blocks.

@@

\subsection{Why is there a GIL?}
The GIL is a very simple synchronisation mechanism for supporting
multithreading in an interpreter. The basic guarantee is that the GIL
may only be released in between bytecode instructions. The interpreter
can thus rely on complete isolation and atomicity of these
instructions. Additionally, it provides the application with a
sequential consistency model~\cite{lamport79}. As a consequence,
applications can rely on certain operations to be atomic and that they
will always be executed in the order in which they appear in the
code. While depending on this may not always be a good idea, it is
done in practice. A GIL-replacement should therefore uphold these
guarantees, while preferably also be as easily implementable as a GIL
for the interpreter. The latter can be especially important since
many of these languages are developed and maintained by very large
open-source communities, which are not easy to coordinate.

@@

\begin{description}
\item[Performance:] How well does the approach perform compared to the
  GIL on a single and on multiple threads?
\item[Existing applications:] How big are the changes required to
  integrate with and parallelise existing applications?
\item[Better synchronisation:] Does the approach enable better, paralleliseable
  synchronisation mechanisms for applications (e.g. atomic blocks)?
\item[Implementation:] How difficult is it to implement the approach
  in the interpreter?
\item[External libraries:] Does the approach allow for easy
  integration of external libraries?
\end{description}

@@

Jython~\cite{webjython} is one project that implements an
interpreter for Python on the Java Virtual Machine (JVM) and
that uses fine-grained locking\footnote{The performance impact of
fine-grained locking is milder in Java than it would be in a typical piece
of C code; see e.g.~\cite{biased}.} to correctly synchronise the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks. Since there is no central location, the
complexity of the implementation is quite a bit larger compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too. One could simply use one lock per library
to avoid this issue.

In the end, fine-grained locking can transparently replace the GIL and
therefore parallelise existing applications, generally without any
changes\footnote{There are rare cases where not having atomic
  bytecodes actually changes the semantics.}. An implementation has to
follow the GIL semantics very closely, otherwise it may expose some
latent data races in existing applications which are just not exposed
with a GIL. This approach does however not provide a better parallelising
synchronisation mechanism to the application like e.g. atomic blocks.

>>>>

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\synctex=-1

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% listings
\usepackage{float}
\floatstyle{ruled}
\newfloat{code}{tbp}{loa}
\providecommand{\codename}{Listing}
\floatname{code}{\protect\codename}

% nice listings
\usepackage{xcolor}
\usepackage{newverbs}

\usepackage{color}
\definecolor{verylightgray}{rgb}{0.93,0.93,0.93}
\definecolor{darkblue}{rgb}{0.2,0.2,0.6}
\definecolor{commentgreen}{rgb}{0.25,0.5,0.37}
\usepackage{letltxmacro}

\usepackage{listings}

\makeatletter
\LetLtxMacro{\oldlstinline}{\lstinline}

\renewcommand\lstinline[1][]{%
  \Collectverb{\@@myverb}%
}

\def\@@myverb#1{%
    \begingroup
    \fboxsep=0.2em
    \colorbox{verylightgray}{\oldlstinline|#1|}%
    \endgroup
}
\makeatother

\lstset{backgroundcolor={\color{verylightgray}},
  basicstyle={\scriptsize\ttfamily},
  commentstyle={\ttfamily\color{commentgreen}},
  keywordstyle={\bfseries\color{darkblue}},
  morecomment={[l]{//}},
  tabsize=4,
  morekeywords={foreach,in,def,type,dynamic,Int,
    Boolean,infer,void,super,if,boolean,int,else,
    while,do,extends,class,assert,for,switch,case,
    private,protected,public,const,final,static,
    interface,new,true,false,null,return}}
\renewcommand{\lstlistingname}{Listing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5

@@

\conferenceinfo{ICOOOLPS'14}{July 28 2014, Uppsala, Sweden}
\copyrightyear{2014}
\copyrightdata{978-1-4503-2914-9/14/07}
\doi{2633301.2633305}

@@

These requirements are not easy to meet. We argue that STM is the
overall winner. While it currently has a big performance problem, it
gets more points in all the other categories. We think that it is the
only solution that also provides a better synchronisation mechanism to
the application in the form of parallelisable atomic blocks.

@@

\subsection{Why is there a GIL?}
The GIL is a very simple synchronisation mechanism for supporting
multithreading in an interpreter. The basic guarantee is that the GIL
may only be released in between bytecode instructions\footnote{This
also applies to Abstract Syntax Tree (AST) interpreters, where the GIL
may only be released between interpreting two AST nodes.}. The interpreter
can thus rely on complete isolation and atomicity for the
instructions' execution. Thus, accesses to data structures like
dictionaries and lists happen atomically and do not need additional
protection from data races when shared between threads.

The GIL also provides the application with a sequential consistency
model~\cite{lamport79}. This can be very valuable as it means less
surprises for the programmer. For example in Table~\ref{tab:seq_cons},
the programmer can expect the critical section to only be entered by
one thread.  If the model allowed to buffer the writes, both threads
may enter the critical section at the same time.

\begin{table}[!ht]
  \begin{center}
    \begin{tabular}{|l|l|}
      \hline
      Thread 1 & Thread 2 \\
      \hline
      \multicolumn{2}{|l|}{\texttt{A = B = 0}} \\
      \hline
      \texttt{A = 1} & \texttt{B = 1}\\
      \texttt{if B == 0:} & \texttt{if A == 0:}\\
      \texttt{    critical section} & \texttt{    critical section}\\
      \hline
    \end{tabular}
    \caption{Critical section with a sequential consistency model.}
    \label{tab:seq_cons}
  \end{center}
\end{table}

As a consequence, applications can rely on certain operations to be
atomic and that they will always be executed in the order in which
they appear in the code. While depending on this may not always be a
good idea, it is done in practice. A GIL-replacement should therefore
uphold these guarantees, while preferably also be as easily
implementable as a GIL for the interpreter. The latter can be
especially important since many of these languages are developed and
maintained by very large open-source communities, which are not easy
to coordinate.

@@

\begin{description}
\item[Performance:] How much does the approach impact performance on a single
  and how much on multiple threads? Can it make use of parallelism?
\item[Existing applications:] How big are the changes required to
  integrate with and parallelise existing applications?
\item[Better synchronisation:] Does the approach enable better,
  paralleliseable synchronisation mechanisms for applications
  (e.g.\ atomic blocks)?  Many synchronisation mechanisms can be built on
  top of all solutions (e.g.\  message passing). We look for mechanisms
  that are directly enabled by the contending approaches.
\item[Implementation:] How difficult is it to implement the approach
  in the interpreter?
\item[External libraries:] Does the approach allow for easy
  integration of external libraries?
\end{description}

@@

Jython~\cite{webjython} is one project that implements an
interpreter for Python on the Java Virtual Machine (JVM) and
that uses fine-grained locking\footnote{The performance impact of
fine-grained locking is milder in Java than it would be in a typical piece
of C code; see e.g.~\cite{biased}.} to correctly synchronise the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks -- every dictionary, list, instance, or mutable
object in general needs a lock. Compared to e.g.\ Java, object
attributes are backed by a dictionary. Accesses to it must be
synchronised because the interpreter could crash otherwise.  Since
there is no central location for all these locks, the
complexity of the implementation is quite a bit larger compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too. One could simply use one lock per library
to avoid this issue.

In the end, fine-grained locking can transparently replace the GIL and
therefore parallelise existing applications, generally without any
changes\footnote{There are rare cases where not having atomic
  bytecodes actually changes the semantics. E.g.\ in Jython,
  \texttt{dict1.update(dict2)} is not atomic: it first reads data from
  \texttt{dict2} with \texttt{dict2}'s lock, and then puts it into
  \texttt{dict1} with \texttt{dict1}'s lock. A lot can happen
  in-between.}. An implementation has to follow the GIL semantics very
closely, otherwise it may expose some latent data races in existing
applications which are just not exposed with a GIL. This approach does
however not provide a better parallelising synchronisation mechanism
to the application like e.g. atomic blocks.
